---
title: "Cluster for microbiome count data via K-means"
author: "Weijia Xiong"
date: "6/25/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(factoextra)
library(gridExtra)
library(corrplot)
library(RColorBrewer) 
library(gplots)
library(tidyverse)
library(Rtsne)
library(vegan)
library(ape)
library(cluster)
```

```{r}
load("data/DiGiulio.RData")
otu_data = as.data.frame(DiGiulio$OTU)  # 927 samples, 1271 OTU
taxonomy = DiGiulio$Taxonomy  # 1271 
sampledata = DiGiulio$SampleData  #  927 samples, other covariates
```

```{r}
otu_data_all= 
  cbind(sampledata, otu_data) %>% 
  mutate(
    Preg = as.factor(Preg),
    Subject = as.factor(Subject)
  ) %>% 
  na.omit()
rownames(otu_data_all) = sampledata$SampleID
```


```{r}
term = 
  otu_data_all %>% 
  filter(preterm == "Term")
  
preterm = 
  otu_data_all %>% 
  filter(preterm != "Term")
```

# Term
```{r}
term_count =
  term %>% 
  select(-SampleID,-Subject,-weeks,-Race,-NumReads,-Preg,-preterm,-CST)
ncol(term_count)

term_filter = term_count[,colSums(term_count) > 0] %>% scale()
ncol(term_filter)
```

## K-means cluster
```{r}
fviz_nbclust(term_filter,
             FUNcluster = kmeans,
             method = "silhouette")

set.seed(1)
km_term <- kmeans(term_filter, centers = 2, nstart = 20)
```

```{r}
km_vis_term <- fviz_cluster(list(data = term_filter, cluster = km_term$cluster), 
                               ellipse.type = "convex", 
                               geom = c("point","text"),
                               labelsize = 5, 
                               palette = "Dark2") + labs(title = "K-means") 

km_vis_term
```


```{r}
term[km_term$cluster == 2, 1:10]
```


## Hierarchical clustering

We can also apply hierarchical clustering on this data. Here we use the Euclidean distance and different types of linkage.

```{r}
dat1 = term_count
hc.complete <- hclust(dist(dat1), method = "complete")

# distance.bray<-vegdist(dat1,method="bray",na.rm=TRUE)          
# hc.bray<- hclust(distance.bray,method="complete")     
```

The function `fviz_dend()` can be applied to visualize the dendrogram.

```{r, fig.width=7}
fviz_dend(hc.complete, k = 4,        
          cex = 0.3, 
          palette = "jco", 
          color_labels_by_k = TRUE,
          rect = TRUE, rect_fill = TRUE, rect_border = "jco",
          labels_track_height = 2.5)
```


```{r}
# Who are in the fourth cluster?
complete <- cutree(hc.complete, 4)
term[complete == 4,1:10]
```

### After scaling and filtering
```{r}
dat1 = term_filter
hc.complete <- hclust(dist(dat1), method = "complete")
fviz_dend(hc.complete, k = 4,        
          cex = 0.3, 
          palette = "jco", 
          color_labels_by_k = TRUE,
          rect = TRUE, rect_fill = TRUE, rect_border = "jco",
          labels_track_height = 2.5)
```

```{r}
complete <- cutree(hc.complete, 4)
preterm[complete == 4,1:10]
```



# Preterm
```{r}
preterm_count =
  preterm %>% 
  select(-SampleID,-Subject,-weeks,-Race,-NumReads,-Preg,-preterm,-CST)
ncol(preterm_count)

preterm_filter = preterm_count[,colSums(preterm_count) > 0] %>% scale()
ncol(preterm_filter)
```

## K-means cluster
```{r}
fviz_nbclust(preterm_filter,
             FUNcluster = kmeans,
             method = "silhouette")

set.seed(1)
km_preterm <- kmeans(preterm_filter, centers = 2, nstart = 20)
```

```{r}
km_vis_preterm <- fviz_cluster(list(data = preterm_filter, cluster = km_preterm$cluster), 
                               ellipse.type = "convex", 
                               geom = c("point","text"),
                               labelsize = 5, 
                               palette = "Dark2") + labs(title = "K-means") 

km_vis_preterm
```


```{r}
preterm[km_preterm$cluster == 1, 1:10]
```


## Hierarchical clustering

We can also apply hierarchical clustering on this data. Here we use the Euclidean distance and different types of linkage.

```{r}
dat1 = preterm_count
hc.complete <- hclust(dist(dat1), method = "complete")

# distance.bray<-vegdist(dat1,method="bray",na.rm=TRUE)          
# hc.bray<- hclust(distance.bray,method="complete")     
```

The function `fviz_dend()` can be applied to visualize the dendrogram.

```{r, fig.width=7}
fviz_dend(hc.complete, k = 4,        
          cex = 0.3, 
          palette = "jco", 
          color_labels_by_k = TRUE,
          rect = TRUE, rect_fill = TRUE, rect_border = "jco",
          labels_track_height = 2.5)
```


```{r}
# Who are in the fourth cluster?
complete <- cutree(hc.complete, 4)
preterm[complete == 4,1:10]
```

### After scaling and filtering
```{r}
dat1 = preterm_filter
hc.complete <- hclust(dist(dat1), method = "complete")
fviz_dend(hc.complete, k = 4,        
          cex = 0.3, 
          palette = "jco", 
          color_labels_by_k = TRUE,
          rect = TRUE, rect_fill = TRUE, rect_border = "jco",
          labels_track_height = 2.5)
```


```{r}
complete <- cutree(hc.complete, 4)
preterm[complete == 4,1:10]
```

