---
title: "cluster using phyloseq"
author: "Weijia Xiong"
date: "6/30/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning = FALSE)
#install_github("phyloseq/joey711")
#BiocManager::install("phyloseq")
library("phyloseq"); packageVersion("phyloseq")
library("ggplot2"); packageVersion("ggplot2")
library("cluster"); packageVersion("cluster")
library("igraph"); packageVersion("igraph")
library("markovchain"); packageVersion("markovchain")
library("RColorBrewer")
library("gridExtra")
library("grid")
library("DESeq2"); packageVersion("DESeq2")
```

# Load data
```{r load-data, message=FALSE}
otu_file <- "data/PregnancyClosed15.RData"
load(otu_file)
```

## Transform the data (proportions):
```{r transform-data}
site <- "Vaginal_Swab"
ps <- PSPreg[[site]]
tt <- data.frame(tax_table(ps))
ps <- transform_sample_counts(ps, function(OTU) OTU/sum(OTU))

```
We are not doing differential abundance analysis here, so the proportion transformation is used for exploratory analyses only.


```{r}
summary(sample_data(ps)$Outcome)
ps_preterm = subset_samples(ps, Outcome %in% c("Preterm","VeryPreterm"))
ps_term = subset_samples(ps, Outcome %in% c("Term","Marginal"))
```

# Term data cluster

The vaginal community is dominated by closely related, but functionally distinct, Lactobacillus species. Therefore it is better to use a non-phylogenetically aware distance measure so as to be able to separate these species. Start with an MDS (or PCoA) ordination:
```{r MDS, warning=FALSE}
braydist <- phyloseq::distance(ps_term, method="bray")
ord = ordinate(ps, method = "MDS", distance = braydist)  
## based in some fashion on the abundance table ultimately stored as a contingency matrix (otu_table-class).

# MDS: Performs principal coordinate analysis (also called principle coordinate decomposition, multidimensional scaling (MDS), or classical scaling) of a distance matrix (Gower 1966), including two correction methods for negative eigenvalues. 

# Need a distance matrix, here use bray-curtis disctance

plot_scree(ord) + xlim(as.character(seq(1,12))) + ggtitle("MDS-bray ordination eigenvalues")

# p1 = plot_ordination(ps, ord, type="taxa", color="Phylum", title="taxa")
# print(p1)

evs <- ord$value$Eigenvalues
print(evs[1:20])
print(tail(evs))
```

## Denoise distance matrix

We would like to clean some of the noise from the data by restricting this to the truly significant dimensions. The top 5 eigenvalues are clearly very significant, but let's keep all the positive eigenvalues that clearly exceed the magnitude of the smallest negative eigenvalues:
```{r PCoA-cutoff2, warning=FALSE}
h_sub5 <- hist(evs[6:length(evs)], 100)
plot(h_sub5$mids, h_sub5$count, log="y", type='h', lwd=10, lend=2)
```


Looks like eigenvalues 6 and 7 still stand out, so we'll go with 7 MDS dimensions.

## Determine number of clusters
We will use the gap statistic to indicate the number of clusters in this data:
```{r gap-stat}
NDIM <- 7
x <- ord$vectors[,1:NDIM]  # rows=sample, cols=MDS axes, entries = value
pamPCoA = function(x, k) {
    list(cluster = pam(x[,1:2], k, cluster.only = TRUE))
}
gs = clusGap(x, FUN = pamPCoA, K.max = 12, B = 50)
plot_clusgap(gs) + scale_x_continuous(breaks=c(seq(0, 12, 2)))
```

The gap statistic strongly suggests at least three clusters, but makes another big jump at K=5 before the slope gets a lot smaller. So, K=5 it is.

## Cluster into CSTs

Perform PAM 5-fold clusters:
```{r pam-k5}
K <- 5
x <- ord$vectors[,1:NDIM]
clust <- as.factor(pam(x, k=K, cluster.only=T))
# SWAPPING THE ASSIGNMENT OF 2 AND 3 TO MATCH RAVEL CST ENUMERATION
clust[clust==2] <- NA
clust[clust==3] <- 2
clust[is.na(clust)] <- 3
sample_data(ps_term)$CST <- clust
CSTs <- as.character(seq(K))
```



## Evaluate clustering

Inspect the results in MDS and NMDS ordinations:
```{r see-pam-k5, message=FALSE}
CSTColors <- brewer.pal(6,"Paired")[c(1,3,2,5,4,6)] # Length 6 for consistency with pre-revision CST+ coloration
names(CSTColors) <- CSTs
CSTColorScale <- scale_colour_manual(name = "CST", values = CSTColors[1:5])
CSTFillScale <- scale_fill_manual(name = "CST", values = CSTColors[1:5])
# grid.arrange(plot_ordination(ps, ord, color="CST") + CSTColorScale,
#              plot_ordination(ps, ord, axes=c(3,4), color="CST") + CSTColorScale, main="Ordination by Cluster")

plot_ordination(ps_term, ord, color="CST") + CSTColorScale

plot_ordination(ps_term, ord, axes=c(3,4), color="CST")  + CSTColorScale

nmds = ordinate(ps_term, method="NMDS", distance=braydist)
plot_NMDS_bray_by_cluster = plot_ordination(ps,nmds, color="CST") + CSTColorScale + ggtitle("NMDS -- bray -- By Cluster")
```


```{r}
sample_data(ps_term)$clust <- clust
samdf <- data.frame(sample_data(ps_term))
table(samdf$clust)
```

