---
title: "Test with DiGiulio data"
author: "Weijia Xiong"
date: "5/11/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(otuSummary)
library(tidyverse)
library(stringr)
```

Hi Weijia, Can you provide a summary of the real data that you explored, including number of patients, how many time points for each of them, are they balanced, etc? Basically, some exploratory data summary/analysis.

DiGiulio’s Vaginal microbiome data is from 40 women. There are 927 samples and 1271 OTU. Each woman has different observation weeks so it is not balanced.

## Load dataset
```{r}
load("data/DiGiulio.RData")
otu_data = as.data.frame(DiGiulio$OTU)  # 927 samples, 1271 OTU
taxonomy = DiGiulio$Taxonomy  # 1271 
sampledata = DiGiulio$SampleData  #  927 samples, other covariates
```

## Summarize the community structure and abundance with OTU table

Using otuReport from otuSummary package. 

```{r}
##combine with taxonomy
taxonomy = 
  taxonomy %>% 
  unite(taxon, Kingdom:Species, sep = ";", remove = FALSE)
otu_all = data.frame(t(otu_data), 
                     taxonomy = taxonomy$taxon)  
# 927 column samples + one taxonomy 
# 1271 OTU rows
```

**specify the taxonomic level: genus**
```{r}
## Using otuReport from otuSummary package
result = otuReport(otutab = otu_all, siteInCol = TRUE, taxhead = "taxonomy",  platform = "qiime", pattern = ";", percent = FALSE, taxlevel = "genus", collap = ";")
```


```{r}
## Filter 10% genus
genus_total = result$readSum
keep_genus = names(genus_total)[genus_total > quantile(genus_total,0.9)]
keep_genus

genus_reads= 
  as.data.frame(
  result$reads,
  row.names = rownames(result$reads)
  )
genus_reads$total = as.numeric(genus_total)

genus_reads_filter = genus_reads[which(row.names(genus_reads) %in% keep_genus),]

dim(genus_reads_filter)

```


## Gain the abundance dataset
```{r}
genus_abundance = as.data.frame(result$Relabund)
rownames(genus_abundance) = rownames(result$reads)
genus_abundance_filter = genus_abundance[which(row.names(genus_abundance) %in% keep_genus),]
dim(genus_abundance_filter)
```

The number of columns: 927, which represents 927 samples.
The number of rows: 269, which represents 269 genus.

After filter, there remains 27 bacteria.


## Combine with sample information
```{r}
genus_abundance_filter_dat = data.frame(t(genus_abundance_filter))

otu_covariate_all= 
  cbind(sampledata, genus_abundance_filter_dat)

head(otu_covariate_all[,1:9])

colnames(otu_covariate_all) = replace(colnames(otu_covariate_all), 9:35,
        sapply(1:27,function(x){str_c("genus_",x)}))
```

Here each row represent one sample.


```{r}
baseline_ID = 
  otu_covariate_all %>% 
  group_by(Subject) %>% 
  summarise(baseline = first(SampleID))

baseline_data = 
  otu_covariate_all %>% 
  filter(SampleID %in% baseline_ID$baseline)

visit_data = 
  otu_covariate_all %>% 
  filter(!SampleID %in% baseline_ID$baseline)
```

```{r}
head(baseline_data[,1:11])
head(visit_data[,1:11])
```

Not the same weeks for each subject.



```{r}
test_genus =
  otu_covariate_all %>% 
  pivot_longer(
    genus_1:genus_5,
    names_to = "genus_type",
    names_prefix = "genus_",
    values_to = "abundance"
  )

test_genus %>% 
  ggplot(aes(x = weeks, y = abundance)) +
  geom_point(aes(x = weeks, y = abundance, color = genus_type),alpha = 0.5) +
  geom_line(aes(x = weeks, y = abundance, color = genus_type))
  
```


```{r}
save.image(file = "Explore_DiGiulio.RData")
```

